---
title: "Packet 4 - Conditional Probability"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Todd CadwalladerOlsker"
date: "*Last updated:* `r Sys.Date()`"
output:
  rmdformats::downcute:
    downcute_theme: "chaos"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, eval = FALSE)
library(rmdformats)
library(openintro)
library(tidyverse)
```

# Probability

Lightly read through Chapter 1 of *Statistical Inference for Everyone*. 

Some questions to start us off: 

1. What is the probability that a single coin flip will land on H?
1. If we flip a coin 10 times, how many times would you expect it to land on H?
1. If we flip a coin 100 times, how many times would you expect it to land on H? 1000 times?
1. If you flip a coin 6 times, and it lands on H every time, what would you expect to happen on the 7th flip?
1. If you flip a coin 4 times, and it lands on H every time, how many heads would you expect out of 10 flips total? (That is, we will flip 6 more times.)
1. If you flip a coin 20 times, and it lands on H every time, how many heads would you expect out of 100 flips total? (That is, we will flip 80 more times.)


## Weather

 If the probability that it will rain next Saturday is 0.25 and the probability that it will rain next Sunday is 0.25, what is the probability that it will rain during the weekend?
 
 1. What is the probability that it rains on either day if we assume that the weather on Saturday and Sunday are **independent**?
 1. What if we suppose that the probabilities are not independent? For example, suppose P(rain on Saturday)=0.25, but
 P(rain on Sunday|rain on Saturday)=0.40 and 
 P(rain on Sunday|no rain on Saturday)=0.20.
 
 What is the probability of rain on Sunday? What is the probability that it will rain during the weekend? Before computing the answer to this last question, make a guess - will your answer be higher or lower than your answer when the weather was independent?


## COVID Rapid Tests

Accoding to [this link](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8530784/), COVID antigen tests (so called "rapid tests") for asymptomatic individuals have a sensitivity of 63.1%, and a specificity of 99.1%. In other words:

* If an individual has COVID, the test will return a positive result 63.1% of the time. 
* If the individual does not have COVID, the test will return a negative result 99.1% of the time.

Some questions before we proceed:

1. Do these tests seem reliable enough to be useful?
1. What percentage of individuals without COVID receive a **false positive** result?
1. I had originally written, "What percentage of positive results are **false positives**?" Why is this question more difficult to answer?
1. What percentage of individuals with COVID receive a **false negative** result?
1. Assuming you are equally likely to actually have COVID or not have COVID, what does a positive test result tell you? What about a negative test result?
1. Why did I say "Assuming..." in the previous question?
1. How well does the quote below from a [recent article](https://www.vox.com/recode/22870328/how-to-use-rapid-tests-omicron) fit with the information we have so far?

"So you might be wondering: What’s the point if rapid tests aren’t as accurate as PCR tests? Well, rapid antigen tests, which look for a specific protein on the Covid-19 virus, remain extremely effective at confirming positive cases. Put simply, if you test positive on a rapid test, you almost certainly have Covid-19. If you test negative, in some cases, you might still test positive on a PCR test, which is much more sensitive because it tests for genetic evidence of the virus. Rapid tests may not pick up positive cases in people who have been vaccinated or who have recently recovered from Covid-19, since they may produce less virus, one expert told Recode."

### A hypothetical population

Imagine we have a population of 10,000 people, and 1% of them (100 people) actually have COVID. We give them all rapid-tests for COVID, and we get the following hypothetical data (based on the sensitivity and specificity values above):

| Number of individuals | Negative Test | Positive Test | Total |
|-----------------------|---------------|---------------|-------|
| No COVID              | 9811          | 89            | 9900  |
| Has COVID             | 37            | 63            | 100   |
| Total                 | 9848          | 152           | 10000 |

1. What is the probability that an individual has COVID, given that they receive a positive rapid-test?
1. What is the probability that an individual does not have COVID, given that they receive a negative rapid-test?
1. How do all of these numbers change if 5% of the population has covid? 10%? 50%? More? 
1. We don't have to use hypothetical data; we can use real data! The first article contains the following data for asymptomatic individuals:

| Number of individuals | Negative Test | Positive Test | Total |
|-----------------------|---------------|---------------|-------|
| No COVID              | 2594          | 11            | 2605  |
| Has COVID             | 38            | 65            | 103   |
| Total                 | 2632          | 76            | 2708  |

According to this data, what is the probability that an individual has COVID, given that they receive a positive rapid-test? 


## The Monty Hall Problem
 
 One of the most popular probability problems is called the Monty
Hall problem, and is based on the television game show "Let’s Make
a Deal." It can take on many forms, but a common form is as follows:

Suppose you’re on a game show, and you’re given the choice
of three doors: behind one door is a car; behind the others, goats. You pick a
door, say No. 1 (but the door is not opened), and the host, who knows what’s
behind the doors, opens another door, say No. 3, which has a goat.^[It's important to note that this is part of the schtick- the host will always open on the the doors you have not chosen, and will always reveal a goat behind the door he opens.] He then says to you, "Do you want to change your choice to door No. 2?" Is it to
your advantage or disadvantage to switch your choice, or does it matter
whether you switch your choice or not?
 
1. What is your initial guess to this problem?
1. What is P(the car is behind Door 1)?
1. What is P(the car is behind Door 1|a goat is revealed behind Door 3)?^[For the last two questions, Door 1 is the door you originally chose, and the host opened door 3, revealing a goat. Door 2 is the door you may switch to.]
1. What is P(the car is behind Door 2|a goat is revealed behind Door 3)?

Need more convincing? Copy the following code into R:
(This code and more can be found at https://github.com/tcadwallclassroom/math-583-packet-4.)

```{r monty_code,eval=FALSE}
## WARNING: This clears all objects out of your memory.
## This can be a good thing.
rm(list=ls())

monty<-function(strat='stay',N=1000,print_games=TRUE)
{
  doors<-1:3 #initialize the doors behind one of which is a good prize
  win<-0 #to keep track of number of wins
  for(i in 1:N)
  {
    prize <- sample(doors,1) #randomize which door has the good prize
    guess <- sample(doors,1) #guess a door at random
    ## Reveal one of the doors you didn't pick which has a bum prize
    if(prize != guess)
      reveal <- doors[-c(prize,guess)]
    else
      reveal<-sample(doors[-c(prize,guess)],1)
    ## Stay with your initial guess or switch
    if(strat=='switch')
      select<-doors[-c(reveal,guess)]
    if(strat=='stay')
      select<-guess
    if(strat=='random')
      select<-sample(doors[-reveal],1)
    ## Count up your wins
    if(select==prize)
    {
      win<-win+1
      outcome<-'Winner!'
    }else
      outcome<-'Loser!'
    if(print_games)
      cat(paste('Guess: ',guess,
                '\nRevealed: ',reveal,
                '\nSelection: ',select,
                '\nPrize door: ',prize,
                '\n',outcome,'\n\n',sep=''))
  }
  #Print the win percentage of your strategy
  cat(paste('Using the ',strat,' strategy, 
            your win percentage was ',win/N*100,'%\n',sep='')) 
}

## Code from: 
## https://www.r-bloggers.com/2012/02/monty-hall-by-simulation-in-r/
```

This creates a function called 'monty' that simulates several runs of the Monty Hall game. 

Try running the following:
```{r monty_ex, eval=FALSE}
monty(strat="stay",N=10)
monty(strat="switch",N=10)
monty(strat="stay",1000,FALSE)
monty(strat="switch",1000,FALSE)
```

## Tuesday's Pet

Here is an interesting variant of a similar idea:

My sister-in-law Laurie loves animals; especially cats and dogs. When she adopts an animal from the shelter, she can never decide between a cat or a dog, so she flips a fair coin -- the probability that her newly adopted pet is a cat or a dog is is exactly 50%/50%, and these probabilities are independent for each pet she adopts. She has adopted two pets at different times over the past year.

1. Laurie (truthfully) tells you that the first pet she adopted is a dog. What is the probability that the second pet is also a dog?
1. Laurie (truthfully) tells you that one of the pets she adopted is a dog. What is the probability that the other pet is also a dog?
1. Laurie (truthfully) tells you that one of the pets is a is a dog adopted on a Tuesday. What is the probability that the other pet is also a dog?

## To Turn In

This assignment is based on two of the labs from Math 338. The setup: 

*"After their Halloween party, Tom and Jessy ended up with 20 candy bars in a bag; Tom thought they had 12 Twix and 8 Reese's candy bars while Jessy thought they had 6 Twix and 14 Reese's."*

The following tables shows what a fair distribution would look like for each possible scenario:

The Fair Distribution of Candies if Tom is Correct:

| |Twix|Reese's|Total|
|---|---|---|---|
|Tom|6|4|10|
|Jessy|6|4|10|
|Total|12|8|20|
 

The Fair Distribution of Candies if Jessy is Correct

| |Twix|Reese's|Total|
|---|---|---|---|
|Tom|3|7|10|
|Jessy|3|7|10|
|Total|6|14|20|

Before any kind of evidence is presented, both believe themselves to be correct, and we have no way of knowing which is more likely to be correct. In other words, our *prior probabilities* for each person being correct is 0.5 (assuming that one of them is, in fact, correct!)

Suppose that they randomly choose 10 candies from the bag, and find 5 Twix and 5 Reese's. 

1. **Without doing any additional calculation,** based solely on the fact that we found 5 Twix and 5 Reese's when we selected 10 candies from the bag, which person do you believe is more likely to be correct?

2. **Without doing any additional calculation,** 
a. Estimate how likely Tom is to be correct based on this new evidence, and estimate how likely Jessy is to be correct, assuming one of them is correct.
b. What if the evidence had been that we pulled out 7 Twix and 3 Reece's? How would that change your answer?


*(Yes, in this example, it would be easy to just pull all 20 candies out of the bag, and determine once and for all who is correct. However, if there were 20,000 candies in the bag, we would not be able to determine who is correct so easily!)*

3. In R, create a distribution of 20 candies based on Tom's claim using the following code:

`Candy_Tom <- c(rep("Twix", 12), rep("Reese's", 8))`

and for Jessy's claim:

`Candy_Jessy <- c(rep("Twix", 6), rep("Reese's", 14))`

Now, we'll simulate drawing 10 candies out of the bag *given* each of these claims. For each claim, we'll do 1000 simulations. Run the following code in R:

```{r tom}
library(tidyverse) # if you have not already loaded it

sims <- 1000
samp_size <- 10
sample_df <- data.frame(matrix(ncol = samp_size, nrow = 0))
colnames(sample_df) <- c(1:samp_size)

for(i in 1:sims) {
  candy_sample <- sample(Candy_Tom, samp_size, replace = F)
  sample_df[i,] <- t(candy_sample)
}

sample_df <- sample_df %>% mutate(Twix_count = rowSums(. == "Twix"))


sample_df %>% ggplot(aes(x = Twix_count)) +
  geom_bar(fill = "forestgreen") + 
  geom_vline(aes(xintercept = 5), linetype = "dashed", color = "firebrick") +
  scale_x_continuous(breaks = c(1:10),labels = c(1:10)) +
  labs(title =  paste(sims, "simulations of drawing", samp_size, "candies"),
       subtitle = "assuming Tom's claim is correct.",
       x = "Number of Twix in sample",
       y = "Number of simulations")
```

Looking at this histogram, *approximate* the probability of drawing exactly 5 Twix *under the assumption* that Tom's claim is correct. You can use

`count(Twix_sims_T, Twix_counter == 5)`

find the exact number of simulations in which exactly 5 Twix were drawn, but remember this number (divided by 1000) is still only an approximation of the probability.

4. Modify the above code to get an approximation of the probability of drawing exactly 5 Twix under the assumption that Jessy's claim is correct.

5. Which person is more liekly to be correct, given our evidence?

6. Now, suppose we have not yet collected any evidence (that is, we have not collected a sample of 10 candies yet.) We're about to draw a sample of 10 candies, so let's propose the following rule:

* If our sample contains 5 or more Twix, we will judge Tom to be correct.
* If our sample contains 4 or fewer Twix, we will judge Jessy to be correct.

To measure how good our rule is, what is the probability of the following *errors*?

a. What is the probability that Tom is correct, but our sample supports Jessy's claim? In other words, what is P(4 or fewer Twix | Tom is correct)?
b. What is the probability that Jessy is correct, but our sample supports Tom's claim? In other words, what is P(5 or more Twix | Jessy is correct)?

7. Could we modify the above rule by increasing or decreasing our cutoff value in such a way that both errors would be reduced?

8. Using Bayes' Theorem and the approximate probabilities you found earlier, what is the probability that Tom is correct given the evidence of drawing 5 Twix? Use the prior probability P(Tom is correct) = 0.5. Compare this to your answer to question 2 -- how good was your earlier guess?

9. What is the probability that Tom is correct given the evidence of drawing 5 Twix, if our prior had instead been P(Tom is correct) = 0.1?
